---
title: "The virus vs. the vaccine"
subtitle: ""
summary: ""
authors: []
tags: []
categories: []
date: 2021-01-21T16:58:37Z
lastmod: 2021-01-21T16:58:37Z
featured: false
draft: false
projects: []
mininote: true
output:  hugodown::md_document  
---

One framing for thinking about COVID is as a race between vaccination and infection, since we expect the former to have at least some mitigating effect on the latter. To get a sense for this I plot the cumulative infection estimates from the [MRC Biostatistics Nowcasting Reports](https://www.mrc-bsu.cam.ac.uk/tackling-covid-19/nowcasting-and-forecasting-of-covid-19/) against the vaccination numbers from [PHE](https://coronavirus.data.gov.uk/details/healthcare) (with historical data filled in from [Our World in Data](https://ourworldindata.org/covid-vaccinations)).

```{r message=FALSE}
try(library(tidyverse), silent=TRUE)
library(jsonlite)
library(ggthemes)
library(lubridate)

b = read_csv("vaccinations.csv")
b$y=b$numFirstDose
b$date=lubridate::dmy(b$date)
b$label = "People vaccinated"


a = fromJSON("mrc_biostats.json")
df=tibble(date=a[[1]],y=a[[2]])
df$date=lubridate::ymd(df$date)

subset = df %>% filter(date>"2021-01-25") %>% arrange(date)%>% mutate(cases = y-lag(y))
mod = lm(log10(cases)~yday(date), data=subset)

dates_to_fill = seq(max(df$date),max(b$date),by="day" )


predicted = tibble(date =dates_to_fill)
predicted$cases= 10^predict(mod,predicted)
predicted$label ="People infected (extrapolated)"
predicted$y=cumsum(predicted$cases) + max(subset$y) - 20000


df$label = "People infected"

dataset=  bind_rows(df,b,predicted)
ggplot(dataset %>% filter(label!="People infected (extrapolated)"),aes(x=date,y=y/1000000,color=label))+geom_line( size=1) + geom_line( size=1,data = dataset %>% filter(label=="People infected (extrapolated)"),lty="11")  +labs(x="Date",y="Number of people (millions)",color="Type",title="Infection and vaccination levels in England",caption="Data sources: MRC Biostatistics Unit (cumulative infection estimates)\nOur World In Data & PHE (vaccination figures – at least one dose)")+theme_hc()+
  theme(plot.title = element_text(hjust = 0.5))+scale_color_manual(name="type",values=c("#ef3939","#ef3939","#2f72f2")) 
ggsave("plot23.png",width=7.3,height=4, type = "cairo") 
```

If we literally were to imagine vaccination and infection to be mutually exclusive (which clearly isn't true in either direction), we could plot a graph like this to look at where they might meet.
```{r}


c=b
c$label = "People unvaccinated"
c$y=56000000-c$y
ggplot(bind_rows(df,c),aes(x=date,y=y/1000000,color=label))+geom_line(size=1) +labs(x="Date",y="Number of people (millions)",color="Type",title="Infection and vaccination levels in England",caption="Data sources: MRC Biostatistics Unit (cumulative infection estimates)\nOur World In Data & PHE (vaccination figures – at least one dose)")+theme_hc()+
  theme(plot.title = element_text(hjust = 0.5))


ggsave("plot2.png",width=7.3,height=4, type = "cairo") 

```

```{r}
df=tibble(date=a[[1]],y=a[[2]])
df$date=lubridate::ymd(df$date)
df$daily_incidence = df$y- lag(df$y)
df$type="What happened (from MRC Biostatus Unit model)"

ggplot(df,aes(x=date,y=daily_incidence))+geom_line()

sixweeklevel = 475833
data = tibble(date = lubridate::ymd("2020-03-22")+(1:(10*7)),daily_incidence=sixweeklevel,type="Original plan")
data2 = tibble(date = lubridate::ymd("2020-03-22")+((10+(10*7):(10+(10*7)+100))),daily_incidence=0,type="Original plan")
data3 = tibble(date = lubridate::ymd("2020-03-22")+(-100:-10),daily_incidence=0,type="Original plan")

ggplot(bind_rows(data,data2,data3,df),aes(x=date,y=daily_incidence,color=type))+geom_line()+theme_classic()+scale_y_continuous(labels = scales::unit_format(unit = "K", scale = 1e-3),expand = c(0,0))+labs(x="Date",y="Daily infections")+scale_color_manual(values=c("red","blue"))

ggsave("plan_comparison.png",type="cairo",width=7,height=4)

```
