---
title: "Five lines of R"
subtitle: ""
summary: ""
authors: []
tags: []
categories: []
date: 2021-03-30T10:18:37Z
featured: false
draft: false
projects: []
mininote: true
output: 
  hugodown::md_document:
    fig_width: 6 
    fig_asp: 0.59
---

Somebody on Twitter asked me whether B.1.1.7 data from Florida was still compatible with a logistic growth curve.

It's amazing how simple this sort of thing is to look at with the Tidyverse and nicely formatted data SGTF data from [Helix](https://github.com/myhelix/helix-covid19db). 
```{r message=FALSE, warning=FALSE,fig.width=7,fig.height=7}
library(tidyverse)
data = read_csv(url("https://raw.githubusercontent.com/myhelix/helix-covid19db/master/counts_by_state.csv"))
state_selection = (data %>% group_by(state) %>% summarise(total=sum(positive)) %>% filter(total>5000))$state
data = data  %>% mutate(percent_sgtf=all_SGTF/positive) %>% filter(state %in% state_selection)
ggplot(data,aes(x=collection_date, y=percent_sgtf))+geom_point()+ stat_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE,  fullrange=TRUE) +xlim(lubridate::ymd("2020-12-01"),lubridate::ymd("2021-04-30"))+labs(title="US SGTF",x="Date",y="Percent SGTF")+facet_wrap(~state)+theme_bw()+scale_y_continuous(label=scales::percent)
```

```{r include=FALSE, echo=FALSE}


ggsave("sgtf_us.png",width=7,height=7,type="cairo")
```

There are lots of ways one could improve this, bringing in genome data and modelling uncertainty, but it provides a quick look at what's happening.
